---
title: "Responsible use of the CVR functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Responsible use of the CVR functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
library(daos)
```

The `cvr_*` functions fetch published annual reports from
Erhvervsstyrelsens distribution service. The service is official, free,
and built for system-to-system access, and annual reports are public
documents by law. Using it is legitimate. Using it *well* is a matter of
a few habits, which is what this article is about.

## The pipeline

Four functions, one step each: build the query, send it, extract the
hits, download the documents. Requires `curl` and `jsonlite`.

```{r}
cvr_query(
  cvr          = c("12345678", "87654321"),
  enddate_from = "2024-01-01",
  enddate_to   = "2024-12-31"
) |>
  cvr_search(contact = "you@example.com") |>
  cvr_hits() |>
  cvr_download("data/pdf")
```

- `cvr_query()` builds the Elasticsearch query (pure, no network).
  Matches by CVR number (max 1000 per query) and accounting-period end
  date.
- `cvr_search()` posts the query and returns the parsed response.
- `cvr_hits()` flattens the response into a tibble with one row per
  document (`cvrnummer`, `dokumenturl`, ...).
- `cvr_download()` downloads the documents, named by CVR number.

Everything before `cvr_search()` is local; that line is the first real
API call.

## Identify yourself

The `contact` argument is mandatory and is sent as the `User-Agent`
header. Erhvervsstyrelsen asks requests to identify the sender, so use a
real email address you can be reached at. The function refuses to run
with an empty contact, but it cannot check that the address is honest;
that part is yours.

## Be polite about volume

Each hit is one published annual report, and its `dokumenter` array
typically holds several files (PDF, XML/XBRL, sometimes XHTML), so
expect roughly 3-5 document rows per company per year in the hits
tibble.

The query selects *companies and periods*. Selecting which *documents*
to download is done in R, between `cvr_hits()` and `cvr_download()`:

```{r}
hits <- hits |> dplyr::filter(dokumentmimetype == "application/pdf")
```

The arithmetic matters: 1000 companies for one year is ~1000 hits but
3000-5000 documents, which at the default `sleep = 3` is 2.5-4 hours of
downloading. Filtering to one mimetype first cuts it to ~1000 files
(under an hour). Always look at the hits tibble before you call
`cvr_download()`; it tells you exactly what you are about to fetch.

`cvr_download()` enforces a delay of at least 1 second between requests
(default 3) and guards existing files with an `overwrite` argument, so
an accidental rerun stops before re-downloading anything.

## Complete results: `scroll`

A single response holds at most `query$size` hits (default 2999).
`cvr_search()` warns if the response is full, since the result may then
be truncated. For large pulls (many companies across many years), use
the scroll API to fetch every matching hit in batches:

```{r}
response <- cvr_query(many_cvrs, "2015-01-01", "2024-12-31") |>
  cvr_search(contact = "you@example.com", scroll = TRUE)
```

`scroll` affects the *completeness of the search result* only. It does
not reduce the number of documents you download afterwards.

## Test small first

Before a big pull, run the pipeline on a handful of companies and a
narrow date range, inspect the hits tibble, and only then scale up. If
you want to see the scroll mechanics work without load, force small
batches: `scroll = 2` on a query with a few hits walks through the whole
flow in a couple of requests.

## Working with the XBRL files

The PDF is the human-readable version of the annual report. For data
work, the XML file next to it is usually the better download: it is
XBRL, the machine-readable filing, and `read_xbrl()` parses it into a
tidy tibble with one row per reported fact. Filter the hits to the XML
mimetype, download, and bind the parsed files with the CVR number as id:

```{r}
hits_xml <- hits |> dplyr::filter(dokumentmimetype == "application/xml")
cvr_download(hits_xml, "data/xml")

xbrl <- list.files("data/xml", full.names = TRUE) |>
  read_files(reader = read_xbrl, out = "bind", .id = "cvr")
```

Two things to know about the result. The `fact` column is text and has
to be converted for numeric work. And a filing carries *both* accounting
years, so filter on the context dates (`startdate`, `enddate`,
`instant`) before comparing companies.

One quirk of XBRL is that amounts often arrive without a meaningful
sign: costs can be reported as positive numbers, and the convention
varies between preparers. `find_signs()` can recover the signs, because
the income statement has to reconcile. It searches for the `+1`/`-1`
assignment that makes the line items sum to the reported total:

```{r}
resultatopgoerelse <- xbrl |>
  dplyr::filter(elementid %in% c(
    "Revenue", "ExternalExpenses", "EmployeeBenefitsExpense",
    "TaxExpense", "ProfitLoss"
  )) |>
  dplyr::mutate(fact = as.numeric(fact))

find_signs(resultatopgoerelse, elementid, fact, total_label = "ProfitLoss")
```

This only works when the set of line items is complete: every component
of the income statement must be there, including the reported result
(`ProfitLoss`, årets resultat). With items missing, the reconciliation
either fails or lands on a wrong assignment. The element names depend on
the taxonomy the company filed under, so start by looking at the
`elementid` values in your own data before hardcoding a selection.

## The legal frame

Annual reports are public by law, and the distribution service is the
channel Erhvervsstyrelsen provides for exactly this kind of retrieval.
Usage falls under the general terms for Danish open public data. The one
substantive restriction to know about is *reklamebeskyttelse* (CVR-lovens
paragraf 19): data on protected companies must not be used for direct
marketing. For statistical and analytical use of financial statements,
that restriction is irrelevant, but if your use case ever drifts toward
contacting the companies, check it.
