---
title: "Getting started with daos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with daos}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(daos)
library(dplyr, warn.conflicts = FALSE)
```

The **daos** package is a collection of utility functions built for
day-to-day statistical work. This vignette walks through every function
with short, self-contained examples.

---

## Convenience shorthands

### `f()`: string interpolation

`f()` is a short alias for `glue::glue()`, inspired by Python's f-string
syntax. The name is not self-explanatory, but it is fast to type and
stays out of the way in a pipeline.

```{r}
year <- 2026
f("Report for {year}")
f("1 + 1 = {1 + 1}")
```

### `nowf()`: formatted timestamp

`nowf()` combines `format()` and `Sys.time()` in a single call, so you
can timestamp an export inline without stopping to look up the
`format(Sys.time(), ...)` signature.

```{r}
nowf()                  # default: YYYYMMDD
nowf("%Y-%m-%d %H:%M")  # custom format
```

A typical use is timestamping an export file:

```{r, eval = FALSE}
iris |>
  writexl::write_xlsx(f("iris_{nowf('%Y%B')}.xlsx"))
```

### `shh()`: suppress messages and warnings

A shorthand for `suppressMessages(suppressWarnings(...))`. Useful when
you are tired of tidyverse startup banners, or when an R script runs as
a CLI tool (e.g. with `rapp`) or in a presentation tool like
`presenterm`, where console output has to be clean. The odd name is on
purpose: a function called `quiet` is one masking conflict away from
trouble.

```{r}
shh(message("this will not appear"))
shh(warning("neither will this"))
```

### `is_blank()`: comprehensive blank test

`is.na()` only catches `NA`, and `rlang::is_empty()` only catches `NULL`
and zero-length vectors. `is_blank()` covers all three plus empty
strings, which is handy when validating input that can arrive in any of
these forms.

```{r}
is_blank(NULL)        # TRUE
is_blank(NA)          # TRUE
is_blank("")          # TRUE
is_blank(c(NA, NA))   # TRUE
is_blank(0)           # FALSE
is_blank("text")      # FALSE
```

### `%??%`: null-coalescing with blank detection

Like `rlang::%||%` but uses `is_blank()` instead of `is.null()`, so it
also triggers on `NA`, `""`, and empty vectors.

```{r}
NULL %??% "default"
NA   %??% 0
""   %??% "unknown"
42   %??% 99          # returns 42
```

### `%like%`: regex matching with NA preservation

`%like%` is a more readable replacement for `str_detect()`.
The infix form reads naturally in a `filter()` call, and unlike
`grepl()`, `NA` values stay `NA` instead of being coerced to `FALSE`.
That matters when filtering on optional fields.

```{r}
c("sedan", "SUV", NA, "truck") %like% "^S"

# In a dplyr pipeline:
filter(ggplot2::mpg, model %like% "\\d+") |>
  count(model)
```

---

## File workflows

### `read_files()`: validate, read, and collect

Reading a set of files normally takes a small pipeline: `list.files()`,
a loop or `purrr::map()`, and manual naming. `read_files()` does it in
one call and handles path expansion, existence checks, format detection,
and naming along the way.

A single path returns the object directly; multiple paths return a named
list with a progress bar.

```{r, eval = FALSE}
# Single file (returns object directly):
df <- read_files("data/results.parquet")

# Multiple files with glue expansion:
lst <- read_files("data/dat{0:9}.parquet", names = 0:9)
```

Supported formats: `csv`, `tsv`, `parquet`, `feather`, `xlsx`, `xls`,
`rds`, `sas7bdat`, `sav`, `por`, `xpt`, `dta`, `json`, `ndjson`,
`jsonl`, `yaml`, `yml`, `txt`.

In practice, `read_files()` works best when data production is
consistent: same formats, same column names, same types across files.
If `read_files()` handles a set of files without warnings, the workflow
is in order. When it warns about type mismatches, something has changed
upstream. There is a separate article about this way of working, and
about using `view_types()` when files refuse to stack:
`vignette("read-files")`.

Supply a custom reader to override auto-detection or add arguments:

```{r, eval = FALSE}
read_files(
  "data/dat{0:9}.parquet",
  reader = \(x) arrow::read_parquet(x, col_select = 1:5)
)
```

**Column name casing**

By default, column names are converted to lowercase after reading
(`.lowercase = TRUE`). Set `.lowercase = FALSE` to preserve original
casing.

**Binding into one tibble**

Set `out = "bind"` to row-bind all files. Use `.id` to track the origin
of each row. When `names` is a numeric vector (e.g. years), the `.id`
column will also be numeric.

```{r, eval = FALSE}
read_files(
  "data/dat{2020:2024}.parquet",
  names = 2020:2024,
  out   = "bind",
  .id   = "year"
)
```

If column types differ across files, a warning is issued and types are
reconciled with `readr::type_convert()`. The `.id` column is always
excluded from this reconciliation.

**Unpacking into individual variables**

Set `out = "unpack"` to assign each file as its own named variable in
the calling environment. By default, `read_files()` aborts if any name
already exists. Set `.overwrite = TRUE` to allow overwriting.

```{r, eval = FALSE}
read_files("data/dat{0:9}.parquet", names = paste0("dat", 0:9), out = "unpack")
```

### `summon()`: retrieve objects by name pattern

A regex-based alternative to `ls()` + `mget()`:

```{r, eval = FALSE}
mget(ls(pattern = "dat\\d"))   # base R
summon("dat\\d")               # daos
```

`summon()` pairs naturally with `out = "unpack"` to collect a family of
objects back into a list:

```{r}
dat1 <- data.frame(x = 1)
dat2 <- data.frame(x = 2)
dat3 <- data.frame(x = 3)

result <- summon("^dat\\d+$")
names(result)
```

### `read_access()`: read from a Microsoft Access database

Connects to an `.mdb` or `.accdb` file via ODBC, executes a SQL query,
and returns the result as a tibble. Requires the `DBI` and `odbc`
packages and a Microsoft Access ODBC driver.

The `verbosity` argument controls output:

- `"compact"` (default): one line per file, best when looping
- `"full"`: header, spinners, and summary, best for single files
- `"quiet"`: no output

```{r, eval = FALSE}
# Full output for interactive use:
df <- read_access("data/sales.mdb", "SELECT * FROM Customers",
                  verbosity = "full")

# Loop over many databases quietly:
files <- list.files("data", pattern = "\\.mdb$", full.names = TRUE)
results <- lapply(files, \(f) read_access(f, "SELECT * FROM Sales"))
```

### `read_xbrl()`: parse an XBRL file

Parses an XBRL XML document and returns a tidy tibble with one row per
fact, joined to context (dates) and unit information. Requires `xml2`.

```{r, eval = FALSE}
df <- read_xbrl("report.xml")
```

The returned tibble has columns: `elementid`, `contextid`, `fact`,
`unitid`, `decimals`, `startdate`, `enddate`, `instant`,
`explicit_member`, `unit`.

For non-UTF-8 files, pass `encoding`:

```{r, eval = FALSE}
df <- read_xbrl("report.xml", encoding = "latin1")
```

The full path from downloaded annual reports to parsed facts, including
sign reconciliation with `find_signs()`, is shown in `vignette("cvr")`.

### Accounts and CVR workflows

Two larger workflows have their own articles instead of a walkthrough
here:

- `vignette("cvr")` covers responsible use of the `cvr_*` pipeline for
  fetching published annual reports from Erhvervsstyrelsens distribution
  service: `cvr_query()`, `cvr_search()`, `cvr_hits()`,
  `cvr_download()`.
- `vignette("accounts")` covers the manual accounts workflow behind
  `accounts_pdf_to_txt()` and `accounts_txt_to_xlsx()`: from PDF
  reports through hand-formatted text files to one validated, tidy
  Excel file.

### `write_excel()` / `append_excel()`: write presentable Excel files

These two functions exist because a plain data export is hard to read
for the people the file is meant for. The complaint from a colleague who
works with macroeconomic statistics was concrete: no frozen header, no
thousand separators, no rounding. `writexl::write_xlsx()` is fast but
writes exactly such unformatted files, and `openxlsx2` can do all the
styling but wants a workbook object, worksheets, styles, and a save
call. `write_excel()` sits in between: one call, with defaults that make
the file presentable.

- Numeric columns with at least one value ≥ 1,000 are formatted with a
  thousand separator and no displayed decimals (`#,##0`); the underlying
  values are preserved.
- Year-like columns are excluded automatically: a numeric column where
  every value is a whole number between 1800 and 2200 is assumed to hold
  years, so `2020` is not displayed as `2.020`. Disable with
  `detect_years = FALSE`, and use `skip_fmt` for columns the heuristic
  cannot guess (e.g. numeric period codes like `202001`).
- `NA` values appear as blank cells.
- The header row is bold.
- The first row is frozen (can be turned off with `freeze_header = FALSE`).

Only `.xlsx` can be written. The old binary `.xls` format is not
supported, and a non-`.xlsx` path fails early instead of producing a
file Excel will complain about.

**Writing a new file**

Pass a data frame or a named list of data frames. Requires
`overwrite = TRUE` if the file already exists.

```{r, eval = FALSE}
# Single data frame: sheet name defaults to "Sheet1"
write_excel(mtcars, "output.xlsx")

# Named list: each element becomes a sheet
write_excel(
  list(Cars = mtcars, Iris = iris),
  "output.xlsx"
)
```

**Sheet naming**

Sheet names come from the list names. Unnamed elements get default names
(`"Sheet1"`, `"Sheet2"`, ...). Mixed naming is also fine:

```{r, eval = FALSE}
# "Hoved" is explicit; second sheet becomes "Sheet2"
write_excel(list(Hoved = mtcars, iris), "output.xlsx")
```

**Appending sheets to an existing file**

`append_excel()` adds sheets without touching the existing content. The
file must already exist, and `overwrite = TRUE` is required if a sheet
of the same name is already present.

```{r, eval = FALSE}
write_excel(list(Hoved = mtcars), "output.xlsx")
append_excel(list(Bilag = iris), "output.xlsx")
```

**Other options**

```{r, eval = FALSE}
# Insert as an Excel table (filter arrows, banded rows)
write_excel(mtcars, "output.xlsx", as_table = TRUE)

# Exclude columns from the #,##0 format
write_excel(mtcars, "output.xlsx", skip_fmt = "hp")
```

### `read_ta()` / `write_ta()`: read and write Greenlandic TA files

```{r, eval = FALSE}
df <- read_ta("ta.file")
write_ta(df, "ta.file")
```

---

## Data inspection

### `view_types()`: compare column types across data frames

`view_types()` is `glimpse()` across several data frames at once: one
row per column, with the type in each dataset. Useful before a join, or
before binding with `read_files(out = "bind")`.

```{r}
df_a <- data.frame(x = 1L,  y = "a", z = TRUE)
df_b <- data.frame(x = 1.0, y = "b", z = 1L)

view_types(df_a, df_b)
```

```{r}
# Only columns where types differ:
view_types(df_a, df_b, diff = TRUE)
```

The `focus` argument checks that a specific column has an expected type
and returns only the datasets where it does not:

```{r}
view_types(df_a, df_b, focus = c(x = "int"))
```

---

## Data validation

The checkpoint pattern behind these two functions has its own article,
`vignette("validation")`, which shows how to build lightweight
validation into pipelines with them.

### `expect_empty()`: assert a data frame is empty

Often you want to assert that a filter finds nothing, and to hear about
it right away when it does. Writing a custom
`if (nrow(x) > 0) cli::cli_abort(...)` every time gets repetitive.
`expect_empty()` does it in one step, fits into a pipeline, and can
write to a log file.

Filter for the rows that should not exist, pipe them in, and the
offending rows are at hand the moment something goes wrong.

```{r}
# Success:
dplyr::starwars |>
  filter(name == "Harry Potter") |>
  expect_empty(
    success_msg = "No Harry Potter rows in starwars",
    warn_msg    = "Unexpected fictional character found"
  )
```

```{r}
# Warning: rows found
dplyr::starwars |>
  filter(mass > 1000) |>
  expect_empty(warn_msg = "Unrealistic mass values")
```

```{r, eval = FALSE}
# Hard abort:
dplyr::starwars |>
  filter(mass > 1000) |>
  expect_empty(abort_msg = "Unrealistic mass values")
```

The `log` argument writes a timestamped entry to a file, so when a
scheduled run misbehaves, the errors are already collected by the time
you sit down to investigate:

```{r, eval = FALSE}
log_path <- f("log/{nowf()}/checks.log")
checker  <- \(data, ...) expect_empty(data, ..., log = log_path)

dplyr::starwars |>
  filter(name == "Harry Potter") |>
  checker(success_msg = "No fictional characters")

dplyr::starwars |>
  filter(height > 250) |>
  checker(warn_msg = "Unusually tall characters")
```

### `flag_duplicates()`: detect and label duplicate rows

Base R's `duplicated()` only marks the second occurrence of a duplicate.
`flag_duplicates()` marks all copies and gives them a shared group ID,
so the whole group can be inspected together. The ID can also serve as
an implicit record linkage key: if several respondents have reported the
same information but background data is missing for one of them, the
shared `dupid` reveals the connection.

```{r}
flag_duplicates(ggplot2::mpg)
```

```{r}
# Check specific columns:
flag_duplicates(ggplot2::mpg, manufacturer, model, year) |>
  filter(isdup) |>
  arrange(dupid)
```

Combine with `expect_empty()` for a pipeline validation gate:

```{r}
ggplot2::mpg |>
  flag_duplicates() |>
  filter(isdup) |>
  expect_empty(warn_msg = "Duplicate rows detected")
```

---

## Data manipulation

### `drop_all_na()`: drop empty rows and/or columns

After a join or import you often end up with rows or columns that are
entirely `NA`. `drop_all_na()` removes them. Unlike `tidyr::drop_na()`
(which drops a row on the *first* `NA`), it only drops a row or column
where *every* value is `NA`. It is equivalent to
`janitor::remove_empty()` with `cutoff = 1`, reimplemented here to keep
the dependency footprint small:

```{r}
df <- tibble::tibble(
  a = c(1, NA, 3),
  b = c(NA, NA, NA),
  c = c("x", NA, "z")
)

drop_all_na(df)                  # column b and the all-NA row
drop_all_na(df, which = "rows")  # only the all-NA row
drop_all_na(df, which = "cols")  # only column b
```

### `split_by()`: split a data frame into a named list

`dplyr::group_split()` returns an unnamed list. `split_by()` adds names
derived from the grouping values, making it straightforward to index by
group:

```{r}
parts <- split_by(ggplot2::mpg, manufacturer)
names(parts)[1:5]
head(parts[["audi"]])
```

Multiple grouping columns are joined with `.sep`:

```{r}
parts2 <- split_by(ggplot2::mpg, cyl, drv, .sep = "-")
names(parts2)[1:4]
```

### `find_signs()`: reconcile accounting line items

`find_signs()` finds the signs (`+1`, `-1`, or `0`) that make a set of
values sum to a given total. Useful when importing accounting data where
the sign conventions are unknown. It uses a *meet-in-the-middle*
algorithm, so it stays fast even with many line items.

```{r}
items <- data.frame(
  label = c("revenue", "costs", "tax", "total"),
  value = c(1000, 400, 100, 500)
)

find_signs(items, label, value, total_label = "total")
```

Pin known signs and allow items to be excluded with `max_zeros`:

```{r, eval = FALSE}
find_signs(
  items,
  label, value,
  total_label = "total",
  positive    = "revenue",
  negative    = c("costs", "tax"),
  max_zeros   = 1L
)
```

---

## Domain-specific

### `clean_cpr()` / `clean_cvr()`: standardise identifier columns

Identifiers rarely arrive in one consistent shape: `111111-1118`,
`12 34 56 78`, `DK12345678`, or a nine-digit CPR that lost its leading
zero in Excel. Before a join, both sides need the same form, and that is
all these two functions do. They take a vector and return a vector, so
the natural place for them is inside a `mutate()`:

```{r}
clean_cpr(c("111111-1118", "101004007", "1111111118"))
clean_cvr(c("DK12345678", "12 34 56 78", "12345678"))
```

The two differ on one point. `clean_cpr()` zero-pads nine-digit values,
because birth days 01 to 09 mean that a third of all CPR numbers start
with a zero, which Excel drops. `clean_cvr()` never adds a digit: a
seven-digit CVR cannot be told apart from a typo, and padding it would
let a malformed number slip past a later `cvr %like% "^\\d{8}$"`
checkpoint.

Neither function validates. Malformed values come back cleaned but still
visibly malformed. Validation is `add_cpr_info()`'s job (below), or a
checkpoint from `vignette("validation")`.

### `dbdot()`: format DB07 industry codes with dots

DB07 industry codes appear both as `011100` and `01.11.00` depending on
the source. `dbdot()` normalises to the dotted form. Existing dots are
stripped first, so mixed input comes out uniform and running it twice
changes nothing. Any aggregation level works:

```{r}
dbdot(c("011100", "01.1100", "0111", "011"))
```

The bare-digit inverse is just `gsub("[.]", "", x)`. Like the `clean_*`
functions, `dbdot()` never invents digits: a code that lost its leading
zero in Excel is ambiguous (`111` could be group `11.1` or class `01.11`),
so keep industry codes as character columns.

### `add_cpr_info()`: add birth date, age, sex, and validity

`add_cpr_info()` appends derived columns to a data frame using official
CPR Register century-detection rules. Dashes and spaces are stripped
automatically, and nine-digit numbers are zero-padded.

A word on validity: `valid` means ten digits encoding a real calendar
date, and nothing more. The modulus-11 check is not part of it, because
the CPR office has assigned numbers without mod-11 control since 2007
and states that they are fully valid. The check is still available as
the separate `mod11` column, where it works as a data quality signal.
The implementation is plain vectorised arithmetic (a digit matrix,
mod-11 as a matrix product, dates built as day counts instead of parsed
strings), so a few million rows take seconds.

```{r}
df <- data.frame(
  pnr = c("1111111118", "111111-1118", "111111118"),
  stringsAsFactors = FALSE
)

add_cpr_info(df, pnr)
```

Choose a subset of columns and optionally rename them:

```{r}
add_cpr_info(df, pnr, add = c(birth_date = "bday", years_old = "age"))
```

A custom reference date shifts the age calculation:

```{r}
add_cpr_info(df, pnr, add = "age", ref_date = "2000-01-01")
```

### `statbank_*`: the Greenland Statbank

The `statbank_*` family is a small client for the Greenland Statbank
(bank.stat.gl). Search the table list, look up a table's variables, and
download the data as a tidy tibble:

```{r, eval = FALSE}
statbank_search("befolkning")

meta <- statbank_meta("BE/BE01/BEXSAT1.PX")
meta$variables

df <- statbank_get(
  "BE/BE01/BEXSAT1.PX",
  tid = c(2023, 2024, 2025),
  art = "Antal"
)
```

Selections in `statbank_get()` are matched against both variable codes
and their Danish display texts, and against both value codes and value
texts, so you can write `art = "Antal"` without knowing that the
internal code is `Number`. Matching also folds Danish letters
(`foedested` matches `fødested`). Variables you do not mention default
to all values.

The result is shaped for programming by default: columns are named by
the variable codes (snake-cased), and cells hold the value codes (so a
sex variable coded 0/1 comes back as 0/1) -- the shape that joins and
models cleanly. Dot-prefixed options change it: `.col_names = "text"`
and/or `.values = "text"` give the display texts instead,
`.values = "both"` keeps the codes and adds a `<column>_txt` column with
the labels alongside, `.clean_names = FALSE` keeps the raw names, and
`.type_convert = FALSE` turns off the automatic `readr::type_convert()`
that otherwise makes e.g. year columns numeric. To reach any PXWeb v1
endpoint, `bank` also accepts a full base URL such as
`"https://bank.stat.gl/api/v1/da/Greenland"`.

The first call to `statbank_search()` or `statbank_tables()` walks the
statbank's table tree (one small request per folder) and caches the
list for the rest of the session.

The same functions also reach the Faroese statbank (Hagstova Føroya) via
`bank = "fo"`. Each bank has its own languages, so `lang` defaults to
the bank's own default (Danish for Greenland, Faroese for the Faroe
Islands) rather than a fixed value:

```{r, eval = FALSE}
statbank_search("wages", bank = "fo")
df <- statbank_get("IP/IP02/pris_alt.px", bank = "fo")
```

`statbank_app()` wraps the family in a Shiny app: find a table by
browsing the subject tree (a three-column, yazi-style browser with
keyboard navigation) or searching, pick values in popup checklists,
preview the data and a plot, and copy the `statbank_get()` call that
reproduces the selection. The bank switcher sits one level above the
subject root, so the same app covers both Greenland (the default) and
the Faroe Islands, and a language chooser switches between each bank's
languages. The app is meant as the bridge from clicking to scripting;
its settings default to the code-first shape (with a "both" option for
`_txt` label columns), and the code tab always shows the query you
built, prefixed `daos::` so it runs without `library(daos)`. Data can be
downloaded as formatted Excel, and pressing Q closes the app and
returns the last fetched dataset to R. Requires `shiny` and `ggplot2`.

```{r, eval = FALSE}
statbank_app()
```

The thinking behind the app -- keyboard-driven, guided navigation of a
statbank, and where that idea is heading -- is its own article,
`vignette("statbank")`.

---

## Interactive tools

### `screen_timeseries()`: interactive time-series screening

It is hard to get as close to data in a console as in Excel, where the
feedback is immediate and visual. `screen_timeseries()` is an attempt to
bridge that gap. It launches a Shiny app where you walk through the time
series group by group, zoom, and flag anomalies -- without writing any
filtering code.

All columns that are not `x`, `y`, `series`, or excluded automatically
become dropdown filters. Requires `shiny`, `ggplot2`, and `plotly`.

```{r, eval = FALSE}
df <- expand.grid(
  year    = 2010:2020,
  country = c("DK", "SE", "NO", "FI")
)
df$gdp <- rnorm(nrow(df), mean = 300, sd = 30)

screen_timeseries(df, x = year, y = gdp)

# Multiple lines per group with series:
ggplot2::economics_long |>
  screen_timeseries(date, value, series = variable)
```

**Keyboard shortcuts**

| Key | Action |
|-----|--------|
| `<-` / `->` | Navigate to previous / next group |
| `Space` | Flag the current group |
| `R` | Reset zoom |
| `Q` | Quit and return flagged groups |

The `.y_min` and `.y_max` arguments fix the Y-axis bounds globally
across all groups. The `.title` argument sets a title shown in the app
header and in downloaded figures:

```{r, eval = FALSE}
screen_timeseries(df, year, gdp,
                  .y_min = 0, .y_max = 500,
                  .title = "GDP by country")
```

### `task_*`: a shared task manager

The `task_*` family is a small taskwarrior-style task manager backed by a
shared SQLite file. SQLite is free (public domain, bundled by `RSQLite`),
runs in WAL mode, and needs no server -- so a `.sqlite` file on a network
drive is enough for a team to work from, several people at once.

```{r, eval = FALSE}
task_db("tasks.sqlite")                       # create/open the database

task_add("tasks.sqlite", "Write the report",
         project = "Q3", assignee = "Anna", tags = c("writing", "urgent"),
         priority = "H", due = "2026-07-01")

task_list("tasks.sqlite")                     # tibble sorted by urgency
task_list("tasks.sqlite", assignee = "Anna")  # filter by person
task_done("tasks.sqlite", id = 1)             # complete a task
```

Tasks carry a project, an assignee (person), tags, priority, due date,
recurrence, dependencies and annotations; `task_list()` adds a `blocked`
flag and a simplified urgency score. Every function opens a short-lived
connection, works in one transaction, and closes, so concurrent users are
handled safely. The database is just a shared file -- point at a path that
does not exist yet and a fresh one is created.

`task_app()` is a Shiny front-end over the same database: add and edit
tasks, filter by person/project/tag/status, see project and people
overviews, and watch the list update as others change it (it re-reads on
a timer). Requires `shiny`, `DBI`, and `RSQLite`.

```{r, eval = FALSE}
task_app("tasks.sqlite")
```

Because every `task_*` function is a plain call against the shared file,
task updates can be embedded in production scripts -- so the same database
the team browses also records how production is going. That way of working
is covered in `vignette("tasks")`.

### `browse_files()`: grab filesystem paths without typing them

Typing or hand-fixing long Windows paths is tedious. `browse_files()`
launches a small Shiny app that walks the filesystem in the same
three-column, yazi-style browser as `statbank_app()` -- parent
directory, current directory, and a live preview -- navigated with
`h`/`j`/`k`/`l` or the arrow keys. The point is to grab paths: mark one
or more files or folders with `Space`, then take them into R.

```{r, eval = FALSE}
p <- browse_files()           # navigate, mark, press Q
```

Pressing `Q` returns the marked paths -- a single string for one, a
`c("a", "b")` vector for several -- with forward slashes, so the result
drops straight into a script. `y` copies the same R expression to the
clipboard, and `o` opens the item under the cursor in the system file
explorer (reusing `open_in_explorer()`). Requires `shiny`.

`browse_files()`, the explorer addins, and `read_files()` together make up
a way of handling many sources without a file explorer open alongside
RStudio; `vignette("paths")` walks through it.

---

## RStudio addins

The following addins are available via the *Addins* menu or the command
palette (`Ctrl+Shift+P`). Keyboard shortcuts can be bound under
*Tools -> Modify Keyboard Shortcuts*.

### Fix Windows paths

Replaces backslashes with forward slashes in Windows-style paths
(`C:\...` or `\\server\...`). Only path-like backslashes are converted;
`\(x)` lambda syntax and escape sequences like `\n` are left alone.

- **Text selected:** operates on the selection only
- **Nothing selected:** operates on the entire active file and restores
  the cursor position

### Text to vector

Converts selected lines (one item per line) to an R character vector
expression. Empty lines are ignored.

Selecting:

```
12345678
87654321
11223344
```

produces:

```r
c(
  "12345678",
  "87654321",
  "11223344"
)
```

Useful for quickly wrapping CVR numbers, variable names, or any
line-delimited list copied from Excel or an email.

### Open in file explorer

Opens a location in the system file explorer.

- **A path selected:** opens it directly if it is a folder, or reveals
  the file inside its containing folder if it is a file
- **An object or call selected:** the selection is evaluated as R code,
  so `my_path` or `file.path(dir, "data.csv")` work too
- **Just the cursor on a token (no selection):** the path-like word under
  the cursor is used, so you can place the cursor on a path or an object
  holding one without selecting it (paths with spaces still need to be
  selected)
- **Cursor on whitespace / nothing selected:** opens the current working
  directory (`getwd()`)

A literal path that exists on disk is used as-is; anything else is
evaluated, making it easy to jump from a path stored in a variable
straight to its location on disk.
