---
title: "Getting started with daos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with daos}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(daos)
library(dplyr, warn.conflicts = FALSE)
```

The **daos** package is a collection of utility functions built for
day-to-day statistical work. This vignette walks through every function
with short, self-contained examples.

---

## Convenience shorthands

### `f()` — string interpolation

`f()` is an alias for `glue::glue()`. It interpolates R expressions
embedded in `{}` inside a string.

```{r}
year <- 2026
f("Report for {year}")
f("1 + 1 = {1 + 1}")
```

### `nowf()` — formatted timestamp

`nowf()` returns the current time as a formatted string. Useful for
building timestamped file names.

```{r}
nowf()                  # default: YYYYMMDD
nowf("%Y-%m-%d %H:%M")  # custom format
```

Combine the two for a quick timestamped path:

```{r, eval = FALSE}
f("log/{nowf()}/pipeline.log")
```

### `quiet()` — suppress messages and warnings

```{r}
quiet(message("this will not appear"))
quiet(warning("neither will this"))

# Useful when loading packages without the startup banner:
# quiet(library(tidyverse))
```

### `is_blank()` — comprehensive blank test

Unlike `rlang::is_empty()`, `is_blank()` treats empty strings and
all-`NA` vectors as blank.

```{r}
is_blank(NULL)        # TRUE
is_blank(NA)          # TRUE
is_blank("")          # TRUE
is_blank(c(NA, NA))   # TRUE
is_blank(0)           # FALSE
is_blank("text")      # FALSE
```

### `%??%` — null-coalescing with blank detection

Returns the right-hand side whenever the left-hand side is blank
(using the same logic as `is_blank()`).

```{r}
NULL %??% "default"
NA   %??% 0
""   %??% "unknown"
42   %??% 99          # returns 42
```

### `%like%` — regex matching with NA preservation

`%like%` wraps `grepl()` but keeps `NA` values as `NA` rather than
converting them to `FALSE`.

```{r}
c("sedan", "SUV", NA, "truck") %like% "^S"

# In a dplyr pipeline:
filter(ggplot2::mpg, model %like% "\\d+") |>
  count(model)
```

---

## File workflows

### `require_files()` — validate paths before reading

`require_files()` expands [glue syntax](https://glue.tidyverse.org/) in
path strings and aborts immediately with a clear message if any file is
missing. It returns a named character vector ready to pipe into
`read_files()`.

```{r, eval = FALSE}
# Expand {0:9} into ten paths and verify they all exist:
require_files("data/dat{0:9}.parquet")
```

### `read_files()` — format-agnostic file reading

`read_files()` detects the format from the extension and dispatches to
the right reader. A single path returns the object directly; multiple
paths return a named list with a progress bar.

```{r, eval = FALSE}
# Single file:
df <- read_files("data/results.parquet")

# Multiple files — returns a named list:
files <- read_files(c("data/a.csv", "data/b.csv"))
```

Supported formats: `csv`, `tsv`, `parquet`, `feather`, `xlsx`, `xls`,
`rds`, `sas7bdat`, `sav`, `por`, `xpt`, `dta`, `json`, `ndjson`,
`jsonl`, `yaml`, `yml`, `txt`.

### `bind_files()` — row-bind a list of data frames

After reading multiple files into a list, `bind_files()` combines them
row-wise. A `source` column (or whatever `.id` names it) records where
each row came from.

```{r}
df1 <- data.frame(year = 2020L, value = 10.5)
df2 <- data.frame(year = 2021L, value = 11.2)
bind_files(list(a = df1, b = df2))
```

When types differ, set `.guess = TRUE` to auto-convert:

```{r, eval = FALSE}
bind_files(list_of_files, .guess = TRUE)
```

### Full pipeline

```{r, eval = FALSE}
require_files("data/dat{0:9}.parquet") |>
  read_files() |>
  bind_files()
```

### `unpack_files()` — assign list elements as variables

Instead of binding, you can unpack a list into individual named objects
in your environment:

```{r, eval = FALSE}
require_files("data/dat{0:9}.parquet") |>
  read_files() |>
  unpack_files()
# dat0, dat1, ..., dat9 are now in your environment
```

### `summon()` — retrieve objects by name pattern

Collect objects whose names match a regex back into a list:

```{r}
dat1 <- data.frame(x = 1)
dat2 <- data.frame(x = 2)
dat3 <- data.frame(x = 3)

result <- summon("^dat\\d+$")
names(result)
```

### `read_ta()` — read Greenlandic TA files

For Greenlandic TA files:

```{r, eval = FALSE}
df <- read_ta("ta.file")
```

---

## Data inspection

### `view_types()` — compare column types across data frames

`view_types()` shows the `pillar::type_sum()` type of each column for
every dataset supplied. Invaluable before joins or before calling
`bind_files()`.

```{r}
df_a <- data.frame(x = 1L,  y = "a", z = TRUE)
df_b <- data.frame(x = 1.0, y = "b", z = 1L)

# All columns:
view_types(df_a, df_b)
```

```{r}
# Only columns where types differ:
view_types(df_a, df_b, diff = TRUE)
```

```{r}
# Check that a specific column has the expected type:
view_types(df_a, df_b, focus = c(x = "int"))
# Returns 0 rows if all match, otherwise shows the offending datasets
```

### `size_env()` — object sizes in an environment

```{r}
big   <- 1:1e6
small <- letters
size_env()        # all objects, largest first
size_env(n = 2)   # top 2 only
```

---

## Data validation

### `expect_empty()` — assert a data frame is empty

Use `expect_empty()` as a checkpoint in a pipeline. It succeeds silently
when the data frame has no rows, warns when it has rows (unless
`abort_msg` is set, in which case it aborts).

```{r}
# Success:
data.frame() |> expect_empty(success_msg = "All good — no rows")

# Warning:
filter(ggplot2::mpg, cyl < 0) |>
  expect_empty(warn_msg = "Unexpected: negative cylinder count")
```

```{r, eval = FALSE}
# Hard error:
filter(dplyr::starwars, height < 0) |>
  expect_empty(abort_msg = "Impossible: negative height")
```

The `log` argument writes a timestamped entry to a file — useful in
automated pipelines where you want a minimal audit trail:

```{r, eval = FALSE}
log_path <- f("log/{nowf()}/checks.log")
checker  <- purrr::partial(expect_empty, log = log_path)

filter(dplyr::starwars, name == "Harry Potter") |>
  checker(success_msg = "No fictional characters")

filter(dplyr::starwars, height > 250) |>
  checker(warn_msg = "Unusually tall characters found")
```

### `flag_duplicates()` — detect and label duplicate rows

`flag_duplicates()` prepends two columns: `isdup` (logical) and `dupid`
(an integer group identifier, `0` for unique rows).

```{r}
flag_duplicates(ggplot2::mpg)
```

```{r}
# Check specific columns:
flag_duplicates(ggplot2::mpg, manufacturer, model, year) |>
  filter(isdup) |>
  arrange(dupid)
```

Combine with `expect_empty()` for a pipeline validation gate:

```{r}
ggplot2::mpg |>
  flag_duplicates() |>
  filter(isdup) |>
  expect_empty(warn_msg = "Duplicate rows detected")
```

---

## Data manipulation

### `split_by()` — split a data frame into a named list

`split_by()` is a named version of `dplyr::group_split()`:

```{r}
parts <- split_by(ggplot2::mpg, manufacturer)
names(parts)[1:5]
head(parts[["audi"]])
```

Multiple grouping columns are joined with `.sep`:

```{r}
parts2 <- split_by(ggplot2::mpg, cyl, drv, .sep = "-")
names(parts2)[1:4]
```

### `find_signs()` — reconcile accounting line items

`find_signs()` uses a *meet-in-the-middle* algorithm to find the sign
assignment (`+1`, `-1`, or `0`) for a set of values such that their
signed sum equals a specified total. Useful when importing accounting
data where sign conventions are unknown.

```{r}
items <- data.frame(
  label = c("revenue", "costs", "tax", "total"),
  value = c(1000, 400, 100, 500)
)

find_signs(items, label, value, total_label = "total")
```

You can pin known signs and allow at most `max_zeros` items to be
excluded:

```{r, eval = FALSE}
find_signs(
  items,
  label, value,
  total_label = "total",
  positive    = "revenue",
  negative    = c("costs", "tax"),
  max_zeros   = 1L
)
```

---

## Danish CPR numbers

### `cpr_info()` — extract birth date, age, sex, and validity

`cpr_info()` appends one or more derived columns to a data frame using
official CPR Register century-detection rules. Dashes and spaces are
stripped automatically; nine-digit numbers are zero-padded to recover
values that lost a leading zero in Excel.

```{r}
df <- data.frame(
  pnr = c("1111111118", "111111-1118", "111111118"),
  stringsAsFactors = FALSE
)

cpr_info(df, pnr)
```

Choose a subset of columns and optionally rename them:

```{r}
cpr_info(df, pnr, add = c(birth_date = "bday", years_old = "age"))
```

A custom reference date shifts the age calculation:

```{r}
cpr_info(df, pnr, add = "age", ref_date = "2000-01-01")
```

---

## Interactive tools

### `track_last_df()` — auto-save the last printed data frame

Enable to automatically capture any data frame returned to the console
as `.last.df` in the global environment. Handy when you forget to assign
an intermediate result.

```{r, eval = FALSE}
track_last_df()          # enable

dplyr::starwars |> head(3)
.last.df                 # the three-row data frame above

track_last_df(FALSE)     # disable
```

### `time2screen()` — Shiny time-series screening dashboard

`time2screen()` launches an interactive Shiny app for reviewing time
series group by group. All columns that are not `x`, `y`, `series`, or
excluded automatically become dropdown filters. Navigate with the
keyboard (`←` / `→`) or click buttons. Press `Esc` to exit.

Requires `shiny` and `highcharter`.

```{r, eval = FALSE}
library(shiny)
library(highcharter)

# Simple example — one line per group:
df <- expand.grid(
  year    = 2010:2020,
  country = c("DK", "SE", "NO", "FI")
)
df$gdp <- rnorm(nrow(df), mean = 300, sd = 30)

time2screen(df, x = year, y = gdp)

# With a series dimension for multiple lines per group:
ggplot2::economics_long |>
  time2screen(date, value, series = variable)
```
