---
title: "Getting started with daos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with daos}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(daos)
library(dplyr, warn.conflicts = FALSE)
```

The **daos** package is a collection of utility functions built for
day-to-day statistical work. This vignette walks through every function
with short, self-contained examples.

---

## Convenience shorthands

### `f()` -- string interpolation

`f()` is a short alias for `glue::glue()`, inspired by Python's f-string
syntax. The name is deliberately brief -- not self-explanatory, but fast
to type and visually unobtrusive in a pipeline.

```{r}
year <- 2026
f("Report for {year}")
f("1 + 1 = {1 + 1}")
```

### `nowf()` -- formatted timestamp

`nowf()` combines `format()` and `Sys.time()` into a single call.
The motivation is simple: when timestamping an export, you want to
write `nowf()` inline without stopping to think about the
`format(Sys.time(), ...)` signature.

```{r}
nowf()                  # default: YYYYMMDD
nowf("%Y-%m-%d %H:%M")  # custom format
```

A common pattern -- timestamping an export file:

```{r, eval = FALSE}
iris |>
  writexl::write_xlsx(f("iris_{nowf('%Y%B')}.xlsx"))
```

### `quiet()` -- suppress messages and warnings

A shorthand for `suppressMessages(suppressWarnings(...))`. Useful when
you grow tired of tidyverse startup banners, or when running R scripts
as CLI tools (e.g. with `rapp`) or in presentation tools like `presenterm`
where console output must be clean.

```{r}
quiet(message("this will not appear"))
quiet(warning("neither will this"))
```

### `is_blank()` -- comprehensive blank test

`is.na()` only catches `NA`. `rlang::is_empty()` only catches `NULL`
and zero-length vectors. `is_blank()` covers all three plus empty
strings -- useful when validating input that may arrive in any of these
forms.

```{r}
is_blank(NULL)        # TRUE
is_blank(NA)          # TRUE
is_blank("")          # TRUE
is_blank(c(NA, NA))   # TRUE
is_blank(0)           # FALSE
is_blank("text")      # FALSE
```

### `%??%` -- null-coalescing with blank detection

Like `rlang::%||%` but uses `is_blank()` instead of `is.null()`, so it
also triggers on `NA`, `""`, and empty vectors.

```{r}
NULL %??% "default"
NA   %??% 0
""   %??% "unknown"
42   %??% 99          # returns 42
```

### `%like%` -- regex matching with NA preservation

`%like%` is intended as a more readable replacement for `str_detect()`.
The infix form reads naturally in a `filter()` call, and unlike `grepl()`,
`NA` values are preserved as `NA` rather than silently coerced to `FALSE`
-- which matters when filtering on optional fields.

```{r}
c("sedan", "SUV", NA, "truck") %like% "^S"

# In a dplyr pipeline:
filter(ggplot2::mpg, model %like% "\\d+") |>
  count(model)
```

---

## File workflows

### `read_files()` -- validate, read, and collect

Reading a set of files in base R or tidyverse requires a pipeline of
`list.files()`, `lapply()` or `purrr::map()`, and manual naming. As a
statistician you want one function for one thing. `read_files()` handles
path expansion, existence checks, format detection, naming, and
collection in a single call.

A single path returns the object directly; multiple paths return a named
list with a progress bar.

```{r, eval = FALSE}
# Single file (returns object directly):
df <- read_files("data/results.parquet")

# Multiple files with glue expansion:
lst <- read_files("data/dat{0:9}.parquet", names = 0:9)
```

Supported formats: `csv`, `tsv`, `parquet`, `feather`, `xlsx`, `xls`,
`rds`, `sas7bdat`, `sav`, `por`, `xpt`, `dta`, `json`, `ndjson`,
`jsonl`, `yaml`, `yml`, `txt`.

In practice, `read_files()` works best when data production is
consistent -- same formats, same column names, same types across files.
If `read_files()` handles a set of files without warnings, the workflow
is in order. When it warns about type mismatches, that is a signal that
something has changed upstream.

Supply a custom reader to override auto-detection or add arguments:

```{r, eval = FALSE}
read_files(
  "data/dat{0:9}.parquet",
  reader = \(x) arrow::read_parquet(x, col_select = 1:5)
)
```

**Column name casing**

By default, column names are converted to lowercase after reading
(`.lowercase = TRUE`). Set `.lowercase = FALSE` to preserve original
casing.

**Binding into one tibble**

Set `out = "bind"` to row-bind all files. Use `.id` to track the origin
of each row. When `names` is a numeric vector (e.g. years), the `.id`
column will also be numeric.

```{r, eval = FALSE}
read_files(
  "data/dat{2020:2024}.parquet",
  names = 2020:2024,
  out   = "bind",
  .id   = "year"
)
```

If column types differ across files, a warning is issued and types are
reconciled with `readr::type_convert()`. The `.id` column is always
excluded from this reconciliation.

**Unpacking into individual variables**

Set `out = "unpack"` to assign each file as its own named variable in
the calling environment. By default, `read_files()` aborts if any name
already exists -- set `.overwrite = TRUE` to allow overwriting.

```{r, eval = FALSE}
read_files("data/dat{0:9}.parquet", names = paste0("dat", 0:9), out = "unpack")
```

### `summon()` -- retrieve objects by name pattern

A regex-based alternative to `ls()` + `mget()`. The base R equivalent
is functional but somewhat clunky:

```{r, eval = FALSE}
mget(ls(pattern = "dat\\d"))   # base R
summon("dat\\d")               # daos
```

`summon()` pairs naturally with `out = "unpack"` to collect a family of
objects back into a list:

```{r}
dat1 <- data.frame(x = 1)
dat2 <- data.frame(x = 2)
dat3 <- data.frame(x = 3)

result <- summon("^dat\\d+$")
names(result)
```

### `read_access()` -- read from a Microsoft Access database

Connects to an `.mdb` or `.accdb` file via ODBC, executes a SQL query,
and returns the result as a tibble. Requires the `DBI` and `odbc`
packages and a Microsoft Access ODBC driver.

The `verbosity` argument controls output:

- `"compact"` (default) -- one line per file, best when looping
- `"full"` -- header, spinners, and summary; best for single files
- `"quiet"` -- no output

```{r, eval = FALSE}
# Full output for interactive use:
df <- read_access("data/sales.mdb", "SELECT * FROM Customers",
                  verbosity = "full")

# Loop over many databases quietly:
files <- list.files("data", pattern = "\\.mdb$", full.names = TRUE)
results <- lapply(files, \(f) read_access(f, "SELECT * FROM Sales"))
```

### `read_xbrl()` -- parse an XBRL file

Parses an XBRL XML document and returns a tidy tibble with one row per
fact, joined to context (dates) and unit information. Requires `xml2`.

```{r, eval = FALSE}
df <- read_xbrl("report.xml")
```

The returned tibble has columns: `elementid`, `contextid`, `fact`,
`unitid`, `decimals`, `startdate`, `enddate`, `instant`,
`explicit_member`, `unit`.

For non-UTF-8 files, pass `encoding`:

```{r, eval = FALSE}
df <- read_xbrl("report.xml", encoding = "latin1")
```

### CVR annual reports -- the download pipeline

Four functions form a pipeline for fetching published annual reports from
Erhvervsstyrelsens distribution service: build the query, send it, extract
the hits, download the documents. Requires `curl` and `jsonlite`.

```{r, eval = FALSE}
cvr_query(
  cvr          = c("12345678", "87654321"),
  enddate_from = "2024-01-01",
  enddate_to   = "2024-12-31"
) |>
  cvr_search(contact = "you@example.com") |>
  cvr_hits() |>
  cvr_download("data/pdf")
```

- `cvr_query()` builds the Elasticsearch query (pure, no network). Matches
  by CVR number (max 1000 per query) and accounting-period end date.
- `cvr_search()` posts the query. The `contact` argument is mandatory and is
  sent as the `User-Agent` header -- Erhvervsstyrelsen requires requests to
  identify the sender. A warning is issued if the result may be truncated
  by the query's `size`; use `scroll = TRUE` to fetch all matching hits in
  batches instead.
- `cvr_hits()` flattens the response into a tibble with one row per
  document (`cvrnummer`, `dokumenturl`, ...).
- `cvr_download()` downloads the documents, named by CVR number, with a
  polite delay between requests (min. 1 second, default 3) and the same
  `overwrite` guard as the other writers in the package.

**Volume -- use the download step with care.** Each hit is one published
annual report, and its `dokumenter` array typically holds several files
(PDF, XML/XBRL, sometimes XHTML), so expect roughly 3-5 document rows per
company per year in the hits tibble. The query selects *companies and
periods*; selecting which *documents* to download is done in R, between
`cvr_hits()` and `cvr_download()`:

```{r, eval = FALSE}
hits <- hits |> dplyr::filter(dokumentmimetype == "application/pdf")
```

The arithmetic matters: 1000 companies for one year is ~1000 hits but
3000--5000 documents, which at the default `sleep = 3` is 2.5--4 hours of
downloading. Filtering to one mimetype first cuts it to ~1000 files
(under an hour).

Note that `scroll` solves a different problem: it is for when the *search
result* holds more hits than the query's `size` allows (e.g. many
companies across many years) -- it does not reduce the number of
documents you download.

The downloaded PDFs feed directly into `accounts_pdf_to_txt()` below.

### `accounts_pdf_to_txt()` -- convert PDF accounts to text

Reads all PDF files in a directory, extracts their text, and writes one
`.txt` file per PDF. Intended as the first step in a manual review
workflow for financial statements. Requires `pdftools`.

```{r, eval = FALSE}
accounts_pdf_to_txt(
  pdf_dir = "data/pdf",
  txt_dir = "data/txt"
)
```

The output directory is created automatically if it does not exist.
File names (typically CVR numbers) are preserved.

### `accounts_txt_to_xlsx()` -- parse accounts text files to Excel

Parses manually formatted text files and exports the result as an Excel
file. Requires `writexl`.

**Text file format**

Each line is either a *category line* or a *data line*:

- **Category line:** a single string with no field delimiter. Sets the
  `note` value for all subsequent data lines.
- **Data line:** three fields separated by `min_spaces` or more
  consecutive spaces: (1) element name, (2) amount for `year`,
  (3) amount for `year - 1`.

Amounts must be whole kroner; periods used as thousands separators are
stripped automatically. If no previous-year amount exists, the third
field may be empty -- it becomes `NA`.

Appending ` statnatio` to a category line negates all amounts in that
category, useful when costs appear with a positive sign in notes. The
suffix is stripped from the final output.

File names are used as identifiers in the `cvr` column. A trailing
`_spec` suffix is stripped automatically.

```{r, eval = FALSE}
df <- accounts_txt_to_xlsx(
  txt_dir  = "data/txt",
  out_file = "data/output.xlsx",
  year     = 2024
)
```

Three validation checks run automatically and abort on failure:

1. Comma in value columns -- indicates wrong decimal separator
2. `NA` in `note` or `elementid` -- indicates a missing category line
3. `NA` in current-year values -- indicates a parsing failure

### `write_pretty_xlsx()` -- write to Excel with sensible defaults

`writexl::write_xlsx()` is fast but bare: no formatting, no frozen header,
no number formatting. `openxlsx2` can do all of that, but its API requires
you to build a workbook object, add worksheets, apply styles, and save --
many lines for what should be a one-liner. `write_pretty_xlsx()` is the
middle ground: a single call with defaults that cover the most common needs.

- Numeric columns with at least one value ≥ 1,000 are formatted with a
  thousand separator and no displayed decimals (`#,##0`); the underlying
  values are preserved.
- `NA` values appear as blank cells.
- The header row is bold.
- The first row is frozen (can be turned off with `freeze_header = FALSE`).

**Writing a new file**

Pass a data frame or a named list of data frames to `data`. Requires
`overwrite = TRUE` if the file already exists.

```{r, eval = FALSE}
# Single data frame -- sheet name defaults to "Sheet1"
write_pretty_xlsx(mtcars, "output.xlsx")

# Named list -- each element becomes a sheet
write_pretty_xlsx(
  list(Cars = mtcars, Iris = iris),
  "output.xlsx"
)
```

**Sheet naming**

Sheet names come from the list names. Unnamed elements get default names
(`"Sheet1"`, `"Sheet2"`, ...). Mixed naming is also fine:

```{r, eval = FALSE}
# "Hoved" is explicit; second sheet becomes "Sheet2"
write_pretty_xlsx(list(Hoved = mtcars, iris), "output.xlsx")
```

**Appending sheets to an existing file**

Use `append` to add sheets without touching the existing content. The file
must already exist, and `overwrite = TRUE` is required if a sheet of the
same name is already present.

```{r, eval = FALSE}
# Add one sheet
write_pretty_xlsx(append = list(Bilag = airquality), path = "output.xlsx")

# Create and append in a single call
write_pretty_xlsx(
  list(Hoved = mtcars),
  "output.xlsx",
  append = list(Bilag = iris)
)
```

**Other options**

```{r, eval = FALSE}
# Insert as an Excel table (filter arrows, banded rows)
write_pretty_xlsx(mtcars, "output.xlsx", as_table = TRUE)

# Exclude columns from the #,##0 format
write_pretty_xlsx(mtcars, "output.xlsx", skip_fmt = "hp")
```

### `read_ta()` / `write_ta()` -- read and write Greenlandic TA files

```{r, eval = FALSE}
df <- read_ta("ta.file")
write_ta(df, "ta.file")
```

---

## Data inspection

### `view_types()` -- compare column types across data frames

Think of `view_types()` as `glimpse()` across multiple data frames at
once -- it shows the type of each column for every dataset supplied.
Invaluable before joins or before binding with `read_files(out = "bind")`.

```{r}
df_a <- data.frame(x = 1L,  y = "a", z = TRUE)
df_b <- data.frame(x = 1.0, y = "b", z = 1L)

view_types(df_a, df_b)
```

```{r}
# Only columns where types differ:
view_types(df_a, df_b, diff = TRUE)
```

The `focus` argument checks that a specific column has an expected type
and returns only the datasets where it does not:

```{r}
view_types(df_a, df_b, focus = c(x = "int"))
```

### `size_env()` -- object sizes in an environment

A simple answer to "what is taking up space?" -- lists all objects in
an environment sorted by size.

```{r}
big   <- 1:1e6
small <- letters
size_env()        # all objects, largest first
size_env(n = 2)   # top 2 only
```

---

## Data validation

### `expect_empty()` -- assert a data frame is empty

Often you want to assert that a filtering condition finds nothing -- and
if it does find something, you want to know immediately with a clear
message. Writing a custom `if (nrow(x) > 0) cli::cli_abort(...)` every
time is repetitive. `expect_empty()` does this in one step, fits into a
pipeline, and optionally writes to a log file.

By filtering for impossibilities and piping through `expect_empty()`,
you have the offending rows at hand the moment something goes wrong.

```{r}
# Success:
dplyr::starwars |>
  filter(name == "Harry Potter") |>
  expect_empty(
    success_msg = "No Harry Potter rows in starwars",
    warn_msg    = "Unexpected fictional character found"
  )
```

```{r}
# Warning -- rows found:
dplyr::starwars |>
  filter(mass > 1000) |>
  expect_empty(warn_msg = "Unrealistic mass values")
```

```{r, eval = FALSE}
# Hard abort:
dplyr::starwars |>
  filter(mass > 1000) |>
  expect_empty(abort_msg = "Unrealistic mass values")
```

The `log` argument writes a timestamped entry to a file. When you
eventually open RStudio to investigate, the errors are already collected:

```{r, eval = FALSE}
log_path <- f("log/{nowf()}/checks.log")
checker  <- \(data, ...) expect_empty(data, ..., log = log_path)

dplyr::starwars |>
  filter(name == "Harry Potter") |>
  checker(success_msg = "No fictional characters")

dplyr::starwars |>
  filter(height > 250) |>
  checker(warn_msg = "Unusually tall characters")
```

### `flag_duplicates()` -- detect and label duplicate rows

Base R's `duplicated()` only marks the second occurrence of a duplicate.
`flag_duplicates()` marks all copies and assigns them a shared group ID
-- making it easy to inspect all instances at once. The ID can also
serve as an implicit record linkage key: if multiple respondents have
reported the same information but background data is missing for one
person, the shared `dupid` reveals the connection.

```{r}
flag_duplicates(ggplot2::mpg)
```

```{r}
# Check specific columns:
flag_duplicates(ggplot2::mpg, manufacturer, model, year) |>
  filter(isdup) |>
  arrange(dupid)
```

Combine with `expect_empty()` for a pipeline validation gate:

```{r}
ggplot2::mpg |>
  flag_duplicates() |>
  filter(isdup) |>
  expect_empty(warn_msg = "Duplicate rows detected")
```

---

## Data manipulation

### `drop_all_na()` -- drop empty rows and/or columns

After a join or import you often end up with rows or columns that are
entirely `NA`. `drop_all_na()` removes them. Unlike `tidyr::drop_na()`
(which drops a row on the *first* `NA`), it only drops a row or column
where *every* value is `NA`:

```{r}
df <- tibble::tibble(
  a = c(1, NA, 3),
  b = c(NA, NA, NA),
  c = c("x", NA, "z")
)

drop_all_na(df)                  # column b and the all-NA row
drop_all_na(df, which = "rows")  # only the all-NA row
drop_all_na(df, which = "cols")  # only column b
```

### `split_by()` -- split a data frame into a named list

`dplyr::group_split()` returns an unnamed list. `split_by()` adds names
derived from the grouping values, making it straightforward to index by
group:

```{r}
parts <- split_by(ggplot2::mpg, manufacturer)
names(parts)[1:5]
head(parts[["audi"]])
```

Multiple grouping columns are joined with `.sep`:

```{r}
parts2 <- split_by(ggplot2::mpg, cyl, drv, .sep = "-")
names(parts2)[1:4]
```

### `find_signs()` -- reconcile accounting line items

`find_signs()` uses a *meet-in-the-middle* algorithm to find the sign
assignment (`+1`, `-1`, or `0`) for a set of values such that their
signed sum equals a specified total. Useful when importing accounting
data where sign conventions are unknown.

```{r}
items <- data.frame(
  label = c("revenue", "costs", "tax", "total"),
  value = c(1000, 400, 100, 500)
)

find_signs(items, label, value, total_label = "total")
```

Pin known signs and allow items to be excluded with `max_zeros`:

```{r, eval = FALSE}
find_signs(
  items,
  label, value,
  total_label = "total",
  positive    = "revenue",
  negative    = c("costs", "tax"),
  max_zeros   = 1L
)
```

---

## Domain-specific

### `cpr_info()` -- extract birth date, age, sex, and validity

`cpr_info()` appends derived columns to a data frame using official
CPR Register century-detection rules. Dashes and spaces are stripped
automatically; nine-digit numbers are zero-padded.

```{r}
df <- data.frame(
  pnr = c("1111111118", "111111-1118", "111111118"),
  stringsAsFactors = FALSE
)

cpr_info(df, pnr)
```

Choose a subset of columns and optionally rename them:

```{r}
cpr_info(df, pnr, add = c(birth_date = "bday", years_old = "age"))
```

A custom reference date shifts the age calculation:

```{r}
cpr_info(df, pnr, add = "age", ref_date = "2000-01-01")
```

---

## Interactive tools

### `track_last_df()` -- auto-save the last printed data frame

Automatically captures any data frame returned to the console as
`.last.df` in the global environment. Handy when you forget to assign
an intermediate result.

```{r, eval = FALSE}
track_last_df()          # enable

dplyr::starwars |> head(3)
.last.df                 # the three-row data frame above

track_last_df(FALSE)     # disable
```

### `screen_timeseries()` -- interactive time-series screening

Many people find it hard to get close enough to data when working in a
console -- Excel's immediate, visual feedback is difficult to replicate.
`screen_timeseries()` is an attempt to bridge that gap: it launches a
Shiny app where you navigate time series group by group, zoom in, and
flag anomalies -- without writing any filtering code.

All columns that are not `x`, `y`, `series`, or excluded automatically
become dropdown filters. Requires `shiny`, `ggplot2`, and `plotly`.

```{r, eval = FALSE}
df <- expand.grid(
  year    = 2010:2020,
  country = c("DK", "SE", "NO", "FI")
)
df$gdp <- rnorm(nrow(df), mean = 300, sd = 30)

screen_timeseries(df, x = year, y = gdp)

# Multiple lines per group with series:
ggplot2::economics_long |>
  screen_timeseries(date, value, series = variable)
```

**Keyboard shortcuts**

| Key | Action |
|-----|--------|
| `<-` / `->` | Navigate to previous / next group |
| `Space` | Flag the current group |
| `R` | Reset zoom |
| `Q` | Quit and return flagged groups |

The `.y_min` and `.y_max` arguments fix the Y-axis bounds globally
across all groups. The `.title` argument sets a title shown in the app
header and in downloaded figures:

```{r, eval = FALSE}
screen_timeseries(df, year, gdp,
                  .y_min = 0, .y_max = 500,
                  .title = "GDP by country")
```

---

## RStudio addins

The following addins are available via the *Addins* menu or the command
palette (`Ctrl+Shift+P`). Keyboard shortcuts can be bound under
*Tools -> Modify Keyboard Shortcuts*.

### Fix Windows paths

Replaces backslashes with forward slashes in Windows-style paths
(`C:\...` or `\\server\...`). Only path-like backslashes are converted
-- `\(x)` lambda syntax and escape sequences like `\n` are left
untouched.

- **Text selected:** operates on the selection only
- **Nothing selected:** operates on the entire active file and restores
  the cursor position

### Text to vector

Converts selected lines (one item per line) to an R character vector
expression. Empty lines are ignored.

Selecting:

```
12345678
87654321
11223344
```

produces:

```r
c(
  "12345678",
  "87654321",
  "11223344"
)
```

Useful for quickly wrapping CVR numbers, variable names, or any
line-delimited list copied from Excel or a mail.

### Open in file explorer

Opens a location in the system file explorer.

- **A path selected:** opens it directly if it is a folder, or reveals
  the file inside its containing folder if it is a file
- **An object or call selected:** the selection is evaluated as R code,
  so `my_path` or `file.path(dir, "data.csv")` work too
- **Just the cursor on a token (no selection):** the path-like word under
  the cursor is used, so you can place the cursor on a path or an object
  holding one without selecting it (paths with spaces still need to be
  selected)
- **Cursor on whitespace / nothing selected:** opens the current working
  directory (`getwd()`)

A literal path that exists on disk is used as-is; anything else is
evaluated, making it easy to jump from a path stored in a variable
straight to its location on disk.
