---
title: "Handling paths and files without a file explorer on the side"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Handling paths and files without a file explorer on the side}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
library(daos)
```

The national accounts run on many sources. Compiling them means reaching
into a lot of folders on a lot of drives -- ministry files, register
extracts, survey results, last year's spreadsheets -- and every one of
them is a path that has to be found, typed correctly, and kept current.
The default way to do this is to keep a file explorer open next to RStudio
and shuttle between the two: locate the file over there, copy the path,
paste it here, fix the backslashes, repeat. It is slow, it breaks
concentration, and it is where a surprising number of "file not found"
errors come from.

Several functions in **daos** exist to close that gap, so that finding,
fixing, and loading paths all happen without leaving the editor. This
article ties them together; each has its own help page.

## The friction is real and specific

Three small annoyances do most of the damage:

1. **Windows paths arrive with backslashes**, which R reads as escape
   characters. Every pasted path needs `\` turned into `/` (or doubled)
   before it is a valid string.
2. **Finding the file means leaving R.** The explorer and the editor are
   two windows, and the path makes a round trip through the clipboard.
3. **Many files at once** -- a dozen sources for one compilation -- turn
   all of the above into a dozen-times chore.

The tools below attack each of these.

## Grab paths without typing them: `browse_files()`

`browse_files()` brings the file explorer *into* R. It launches a small
Shiny app with the same three-column, yazi-style browser as
`statbank_app()` -- parent directory, current directory, and a live
preview -- navigated entirely by keyboard with `h`/`j`/`k`/`l` or the
arrow keys. The point is not to look at files but to grab their paths.

```{r}
p <- browse_files()    # navigate, mark with Space, press Q
```

Mark one or more files or folders with `Space`, then `Q` returns them to
R -- a single string for one, a `c("a", "b")` vector for several -- with
forward slashes already, so the result drops straight into a script with
nothing to fix. `y` copies the same R expression to the clipboard, and `o`
opens the item under the cursor in the system explorer when you do need the
real thing. There is no network and no path round-trip through another
window: you found the file and you have its path, in one place.

## Jump from a path in your code to its folder: `addin_open_in_explorer()`

The reverse direction is just as common: you have a path in your script and
you want to see it on disk -- check that the file is there, that it is the
right version, that the folder has what you expect. `open_in_explorer()`
does it from the console, and the **Open in file explorer** addin does it
from wherever your cursor is.

```{r}
open_in_explorer("//server/share/sources/2026")
open_in_explorer(my_source_dir)   # an object holding a path works too
```

The addin resolves its target intelligently: the selected text if there is
a selection; otherwise the path-like token under the cursor, so you can
just rest the cursor on a path or on a variable holding one; otherwise the
working directory. A literal path is opened as-is, anything else is
evaluated as R code, and a file is *revealed* (selected) inside its folder
rather than opened. So `my_source_dir`, `file.path(root, "data.csv")`, or a
bare path all take you to the right place without selecting and copying
first.

This is also why directory links in the package's `cli` messages are
clickable -- they point at this same `open_in_explorer()`.

## Fix backslashes after a paste

When a path does come in over the clipboard -- from an email, a colleague,
a dialog box -- three tools turn it into valid R, depending on how it
arrived:

- **`addin_paste_path()`** reads the clipboard, flips the backslashes, and
  inserts the result as a quoted string at the cursor. This is the
  one-step "I just copied a path in Explorer" path: paste straight into R,
  already quoted and corrected.
- **`addin_fix_path()`** repairs paths *already in your file*. With a
  selection it fixes the selection; with nothing selected it fixes the
  whole file, touching only genuine Windows paths (`C:\...`, `\\server\...`)
  and leaving `\(x)` lambdas and `\n` escapes alone.
- **`addin_flip_backslash()`** is the blunt version: every `\` in the
  selection becomes `/`, no questions asked, for when you know that is all
  you want.

Bind whichever you reach for most to a keyboard shortcut under *Tools ->
Modify Keyboard Shortcuts*, and the backslash problem stops being a
problem.

## Turn a list of files into an R vector

Sources often arrive as a list -- filenames in an email, codes in a
spreadsheet column, one per line. `addin_text_to_vector()` wraps a
selection of lines into a `c(...)` expression:

```
2026_q1.parquet
2026_q2.parquet
2026_q3.parquet
```

becomes

```r
c(
  "2026_q1.parquet",
  "2026_q2.parquet",
  "2026_q3.parquet"
)
```

so a pasted column becomes a vector you can `read_files()` over, without
hand-quoting each line.

## From paths to data: `read_files()` and `summon()`

Once the paths are in hand, `read_files()` is the step that turns many
sources into many objects -- one call that expands `glue`-style patterns,
checks existence, detects the format, reads, and names the results, with a
progress bar for a batch. It is the natural other end of `browse_files()`:
find the files, then read them.

```{r}
# many quarterly sources at once
read_files("//server/share/sources/2026_q{1:4}.parquet", names = 1:4)

# or bind them into one tibble, tracking the origin
read_files(
  "//server/share/sources/{2015:2024}.parquet",
  names = 2015:2024, out = "bind", .id = "year"
)
```

`summon()` is the bookkeeping companion: a regex over the objects you have
loaded, so a family of sources read as separate variables can be gathered
back into a list without retyping their names.

```{r}
summon("^source_\\d+$")
```

The reading side -- formats, custom readers, and what to do when files
refuse to stack because their types drifted -- has its own article,
`vignette("read-files")`. This one is about the step before that: getting
the paths right, fast, without a second window.

## The shape of the workflow

Put together, a national-accounts compilation stops needing a file explorer
alongside it:

1. `browse_files()` to mark this cycle's sources and drop their paths into
   the script.
2. `addin_paste_path()` / `addin_fix_path()` for the odd path that comes in
   over the clipboard.
3. `read_files()` to load them all in one call.
4. `open_in_explorer()` (or the addin) when you do need to see a file on
   disk -- straight from the path in your code.

Every step stays inside RStudio, which is the whole point: the paths never
make a round trip through another window, so there is nowhere for a
backslash or a stale path to sneak in.
