---
title: "Less fiddling with file paths"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Less fiddling with file paths}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
library(daos)
```

Working with files in R means dealing with paths, and on Windows that is a
small but constant annoyance. Nothing here is hard -- it is just friction:
backslashes that R will not accept, a file explorer you keep switching to
and from, and the same little chores repeated every time you point at a new
file.

This is not a style guide. It is just a few small attempts to make working
with Windows paths a bit less tedious, collected as functions in **daos**.
Take whatever you find useful and ignore the rest.

## The annoyances

Three small things cause most of it:

1. **Pasted Windows paths have backslashes**, which R reads as escape
   characters, so every path needs `\` turned into `/` before it is a valid
   string.
2. **Finding a file means leaving R** for the file explorer, copying the
   path, and pasting it back.
3. **It repeats.** A handful of files in one script means doing all of the
   above a handful of times.

The functions below each take one of these off your plate.

## Grab a path without typing it: `browse_files()`

`browse_files()` is a small file explorer that lives inside R. It opens a
three-column, keyboard-driven browser -- parent folder, current folder, and
a preview -- that you move through with `h`/`j`/`k`/`l` or the arrow keys.

```{r}
p <- browse_files()    # navigate, mark with Space, press Q
```

Mark one or more files or folders with `Space`, then `Q` hands the paths
back to R -- a single string for one, a `c("a", "b")` vector for several --
already with forward slashes, so there is nothing to fix. `y` copies the
same expression to the clipboard, and `o` opens the item under the cursor
in the system explorer. No second window, no clipboard round-trip.

Pass `root` to cap how far up you can climb -- the browser never goes above
that directory, so you can scope it to a project folder:

```{r}
browse_files("data", root = "data")   # cannot navigate above data/
```

`p` toggles the content preview -- the text peek and Excel cell peek -- while
the preview column itself stays put.

Press `r` to toggle *reader mode*: instead of the bare path, `Enter` and `y`
insert a call that reads the target with the native reader for its type --
`arrow::read_parquet()` for `.parquet`, `readxl::read_xlsx()` for `.xlsx`,
`readr::read_csv2()` for `.csv`, and so on. A single file is read inline --

```{r}
arrow::read_parquet("data/x.parquet")
```

while several files of the same type are bound to a `my_paths` object and the
reader is mapped over them, so the vector is named and left around to reuse:

```{r}
my_paths <- c(
  "data/a.parquet",
  "data/b.parquet"
)

lapply(my_paths, arrow::read_parquet)
```

Pass `map = "purrr"` to `browse_files()` to get `purrr::map()` instead of
`lapply()`, or toggle it live in the app with `m`. A mix of types can't share
one reader, so each file gets its own named read instead. The inserted code is self-contained -- it carries the
reader it needs and runs without `daos` attached.

Press `b` (or set `base_dir = TRUE`) to factor a shared directory out into a
`base_dir` variable, so a long common path lives in one place:

```{r}
base_dir <- "C:/Users/you/demo"

my_paths <- c(
  paste0(base_dir, "/a.parquet"),
  paste0(base_dir, "/b.parquet")
)

lapply(my_paths, arrow::read_parquet)
```

## Jump to a file you already have in code: `open_in_explorer()`

The other direction: you have a path in your script and want to see it on
disk. `open_in_explorer()` does it from the console, and the **Open in file
explorer** addin does it from wherever the cursor is.

```{r}
open_in_explorer("C:/data/2026")
open_in_explorer(my_dir)   # a variable holding a path works too
```

The addin figures out its target on its own: the selected text, or the
path-like word under the cursor (so you can just rest the cursor on a path
or an object holding one), or the working directory if there is nothing.
A literal path opens directly; anything else is evaluated as R code; a file
is revealed (selected) inside its folder rather than opened.

## Fix the backslashes

When a path does arrive over the clipboard, three addins turn it into valid
R, depending on how it came in:

- **`addin_paste_path()`** reads the clipboard, flips the backslashes, and
  inserts the result as a quoted string -- the one-step "I just copied a
  path" move.
- **`addin_fix_path()`** repairs paths already in your file: the selection,
  or the whole file if nothing is selected. It only touches real Windows
  paths and leaves `\(x)` lambdas and `\n` escapes alone.
- **`addin_flip_backslash()`** is the blunt version: every `\` in the
  selection becomes `/`, for when that is all you want.

These are RStudio addins, like the explorer one above. I prefer to call
them from the command palette (`Ctrl+Shift+P`, then type the addin name),
but you can also bind whichever you use most to a keyboard shortcut under
*Tools -> Modify Keyboard Shortcuts*.

## Turn a list of filenames into a vector

Filenames often arrive as a list -- one per line in an email or a
spreadsheet column. `addin_text_to_vector()` wraps a selection of lines
into a `c(...)` expression, so

```
jan.csv
feb.csv
mar.csv
```

becomes

```r
c(
  "jan.csv",
  "feb.csv",
  "mar.csv"
)
```

without hand-quoting each line.

## Then read them: `read_files()`

Once the paths are in hand, `read_files()` reads them -- one call that
expands `glue`-style patterns, checks the files exist, detects the format,
and names the results, with a progress bar for a batch.

```{r}
read_files("C:/data/2026_q{1:4}.parquet", names = 1:4)
```

`summon()` is its companion for gathering objects you have already read back
into a list by a name pattern. The reading side -- formats, custom readers,
and binding files that will not stack -- has its own article,
`vignette("read-files")`. This one is just about getting the paths right
with less fuss.
