---
title: "Reading many files at once"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading many files at once}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
library(daos)
library(dplyr, warn.conflicts = FALSE)
```

A surprisingly large share of production scripts starts the same way:
one `read_*()` call per file.

```{r}
dat2020 <- read_parquet("data/dat2020.parquet")
dat2021 <- read_parquet("data/dat2021.parquet")
dat2022 <- read_parquet("data/dat2022.parquet")
dat2023 <- read_parquet("data/dat2023.parquet")

df <- bind_rows(dat2020, dat2021, dat2022, dat2023)
```

Reading files one by one can go well -- and it can go quietly wrong. The
block grows with every new delivery, the calls drift apart as one of
them picks up an extra argument, a missing file is only discovered when
its own line runs, and `bind_rows()` at the end either fails or, worse,
succeeds while hiding a type change. This article is about replacing the
block with one `read_files()` call -- and about using `view_types()`
when the files turn out not to be as stackable as they looked.

## One call, many files

`read_files()` expands paths with glue syntax, checks that *every* file
exists before reading anything, detects the format from the extension,
and returns a named list (with a progress bar along the way):

```{r}
lst <- read_files("data/dat{2020:2023}.parquet", names = 2020:2023)
```

Two things are easy to miss but matter in practice:

- **Missing files fail fast and together.** One error lists every absent
  file up front, instead of the loop dying at file seven after five
  minutes of reading.
- **Column names are lowercased by default**, which removes the most
  common cosmetic difference between deliveries before it becomes a
  problem (`.lowercase = FALSE` to opt out).

## Stacking files that belong together

When the files are slices of the same dataset -- years, regions, batches
-- bind them directly with `out = "bind"`, and use `.id` to keep track
of where each row came from:

```{r}
df <- read_files(
  "data/dat{2020:2023}.parquet",
  names = 2020:2023,
  out   = "bind",
  .id   = "year"
)
```

## When the stack does not stack

Files that *should* be identical often are not: the same column arrives
as `chr` in one year and `dbl` in the next, or a column was renamed
upstream. With the one-call-per-file pattern this surfaces -- if at all
-- as a confusing `bind_rows()` error or a silently mistyped column.

`read_files(out = "bind")` makes it a managed event instead. If the
files cannot be bound as-is, it warns, coerces everything to character,
re-types the result with `readr::type_convert()`, and points you to
`view_types()`:

```
Warning: Column types differ across files -- coercing to character and
re-typing with `readr::type_convert()`.
i Use `daos::view_types` to inspect the differences.
```

The rule of thumb: if `read_files()` handles a set of files without
warnings, the data production is consistent. When it warns, something
changed upstream -- and the next step is to find out what.

## Finding the culprit with `view_types()`

`view_types()` is `glimpse()` across multiple data frames at once: one
row per column, one column per dataset, with abbreviated type strings.

```{r, eval = TRUE}
dat2023 <- tibble(cvr = "11111111", name = "Firm A", amount = 100,    year = 2023L)
dat2024 <- tibble(cvr = 11111111,  name = "Firm B", amount = "1.000", turnover = 200)

view_types(dat2023, dat2024)
```

The full listing already tells the story, but `diff = TRUE` cuts it down
to the columns that actually disagree:

```{r, eval = TRUE}
view_types(dat2023, dat2024, diff = TRUE)
```

Read the output like this:

- A row with two different type strings (`chr` vs `dbl` for `cvr`) is a
  type mismatch -- the classic cause of a failed bind or a corrupted
  join key.
- An `NA` means the column does not exist in that dataset at all
  (`year` vs `turnover` above) -- typically a rename or a dropped
  column upstream.

The same overview is just as useful *before a join*: two frames that are
about to be joined on `cvr` should agree on its type, and `view_types()`
answers that in one line instead of two `glimpse()` calls and a visual
diff.

For pipelines, the `focus` argument turns the inspection into a check:
it returns only the datasets where a critical column does *not* have the
expected type, and an empty result on success -- which plugs straight
into the checkpoint pattern from `vignette("validation")`:

```{r, eval = TRUE}
view_types(dat2023, dat2024, focus = c(amount = "dbl")) |>
  expect_empty(warn_msg = "amount is not numeric everywhere")
```

## Other shapes of output

Not every set of files should be stacked. The default (a named list)
pairs well with `lapply()` or `purrr::map()`, and `out = "unpack"`
assigns each file as its own variable -- closest to the
one-call-per-file pattern, but with the validation, naming, and progress
handled for you:

```{r}
read_files("data/dat{2020:2023}.parquet",
           names = paste0("dat", 2020:2023), out = "unpack")
```

`summon("^dat\\d+$")` collects an unpacked family back into a named
list when you change your mind.
