---
title: "Browsing a statbank by keyboard"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Browsing a statbank by keyboard}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
library(daos)
```

A statbank can be hard to navigate. The subjects sit in collapsible trees,
each table has its own set of variables, and choosing the data means
ticking values across several of them. It works, but it is easy to lose
the thread -- to forget which values you have selected in and which you
have left out.

`statbank_app()` is my attempt at a different feel for it. It is not a
finished product, just a take on what navigating a statbank could be like.

```{r}
statbank_app()           # Greenland (default)
statbank_app(bank = "fo") # the Faroe Islands
```

You drift through it by keyboard. Tables are found in a three-column
browser -- parent folder, current folder, and a live preview of whatever
the cursor is on -- moved with `h`/`j`/`k`/`l` or the arrow keys, so you
can see what a folder holds before you open it. Values are picked in popups
with a search box, select/deselect-all, and a running count, so you always
have an overview of what you are choosing and what you are leaving out. And
the code tab always shows the `statbank_get()` call that reproduces your
selection, so an exploration by hand turns into a script you can keep.

It is not perfect. But as a newcomer I would take this over a wall of
collapsible trees, where it is much harder to keep track of what you have
actually selected.

## Where it could go: guidance while you drift

The piece I would add next is better guidance as you move. Right now the
app shows you *what values exist*; the more useful thing is a hint about
*what they mean*, right next to the checkbox.

Take a value shown in current prices or as chained values. If you want the
volume development over time -- real growth, with price changes stripped
out -- you want chained values, but a label reading `Løbende priser` versus
`Kædede værdier` does not tell a newcomer that. A small note in the popup
could:

> **Chained values** -- pick this for volume development (real growth over
> time).
>
> **Current prices** -- pick this for the nominal size in a given year.

So you are met with the relevant bit of guidance while you browse, instead
of downloading the wrong series and only then noticing. This does not exist
yet, here or in the official interface, but it is the direction the app is
pointing.

## Using the client directly

If you already know the table and the values, the `statbank_*` client is
what the app builds calls to, and you can skip the app entirely:

```{r}
statbank_search("national accounts")
meta <- statbank_meta("NR/NR01/NRXxxx.PX")   # the variables and values

df <- statbank_get("NR/NR01/NRXxxx.PX", tid = 2015:2024,
                   enhed = "Kædede værdier")
```

Selections are matched against both codes and their display texts, so you
can write what you see. By default the result comes back code-first --
snake-cased code column names and coded cells, ready to join and model --
with `.col_names = "text"`, `.values = "text"`, or `.values = "both"` (codes
plus `_txt` label columns) to change that. The full client walkthrough is in
`vignette("daos")` under `statbank_*`.
