---
title: "Browsing a statbank by keyboard"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Browsing a statbank by keyboard}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
library(daos)
```

`statbank_app()` is not really meant as a finished product. It is a take
on a question: what would it feel like to move through a statistics bank
the way you move through a filesystem in a terminal -- by keyboard, fast,
and with the program telling you what exists at each step instead of
making you know it in advance? The `statbank_*` client underneath is a
plain API wrapper; the app is the argument that navigation could be nicer
than a tree of dropdowns.

This article is about that argument -- what the app does today, and where
the idea is pointing.

## The problem with picking from a statbank

A statbank table is a set of variables, each with its own list of values,
and a table is fetched by choosing some values from each. The standard web
interface presents this as nested menus: open a folder, open a table, then
a wall of multi-selects. It works, but it asks you to already know what you
are looking for. You pick `art = Number` without the interface ever
hinting that the table also offers an index, or chained values, or a
share. The information you need to choose well is exactly the information
the menu does not surface while you choose.

## What the app does now

`statbank_app()` is built around being shown what exists as you go.

```{r}
statbank_app()           # Greenland (default)
statbank_app(bank = "fo") # the Faroe Islands
```

**Finding a table** happens in a three-column, yazi-style browser -- parent
folder, current folder, and a live preview of whatever the cursor is on --
the same browser as `browse_files()`. You walk the subject tree with
`h`/`j`/`k`/`l` or the arrow keys, and the preview column tells you what a
folder contains or what a table is before you commit to opening it. Or you
search the titles and jump straight there. The bank chooser sits one level
*above* the subject root, so pressing `h` at the top switches between
Greenland and the Faroe Islands -- a bank is just the parent of everything
else.

**Choosing values** adapts to the variable. Time, and numeric variables
with many values such as age, get from/to dropdowns. Everything else opens
a popup with a searchable checkbox list, select/deselect-all, and a running
count, so a variable with two hundred municipalities is browsable instead
of overwhelming. An empty selection means all values, which keeps the
common case to zero clicks.

**Fetching** shows the data as a table and a plot, lets you download a
formatted Excel file, and -- the part that matters most -- always shows the
`statbank_get()` call that reproduces exactly what you selected. The app is
a bridge from clicking to scripting: you explore by hand once, copy the
query, and from then on it runs headless.

```{r}
# the kind of call the code tab hands you
statbank_get(
  "BE/BE01/BEXSAT1.PX",
  tid = c(2023, 2024, 2025),
  art = "Antal"
)
```

Pressing `Q` closes the app and returns the last fetched dataset, so
`df <- statbank_app()` is itself a way to get data into R.

## Guidance is already half the point

Notice how much of the above is the program telling you what is there. The
preview column tells you what a folder holds. The popup's running count
tells you how big your selection is. The code tab tells you what query you
just built, in a form you can keep. None of that is data you had to bring;
the app surfaces it at the moment you need it. That is the thread worth
pulling on.

## Where this is heading: guidance on what a choice *means*

Today the app guides you on *what values exist*. The more interesting step
is guidance on *what those values mean* -- and that part does not exist
yet, in this app or in the official interface.

Take the national accounts. A unit variable might offer current prices,
chained values, an index, and a few more, and the choice is not cosmetic:
if you want the volume development of an aggregate, you want chained
values, and a newcomer has no way to know that from a label reading
`2020-priser, kædede værdier`. The knowledge exists -- it is in the
methodology, in colleagues' heads -- but never next to the checkbox where
the decision is actually made.

The direction `statbank_app()` is sketching is to put it there. Imagine the
value popup carrying a line of plain-language help per option:

> **Chained values** -- pick this for volume development (real growth over
> time), stripping out price changes.
>
> **Current prices** -- pick this for the nominal size of the economy in a
> given year.

So you are met with the relevant guidance while you browse, not after you
have downloaded the wrong series and noticed it does not behave. The same
idea generalises beyond national accounts: a "what is this and when do I
want it" note attached to the variables and values where the choice is
genuinely consequential.

This is not implemented. It is the reason the app is framed as a proof of
concept rather than a tool: the navigation, the previews, and the code tab
are there to show that a statbank *could* guide you through a selection,
and the value-level guidance is the next thing that idea asks for.

## Using the client directly

If you already know the table and the values, you do not need the app at
all. The client is small and scriptable, and the app only ever produces
calls to it:

```{r}
statbank_search("national accounts")

meta <- statbank_meta("NR/NR01/NRXxxx.PX")
meta$variables          # what the app would show you in the popups

df <- statbank_get(
  "NR/NR01/NRXxxx.PX",
  tid  = 2015:2024,
  type = "Chained values"
)
```

Selections are matched against both codes and their display texts, and
Danish letters fold (`foedested` matches `fødested`), so you can write what
you see. The full client walkthrough is in `vignette("daos")` under
`statbank_*`.
